---
title: "Data_preprocessing_cfRNA"
author: "Yayuan Zhu"
date: "2024-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup}
# When writing the text, add references for these packages.
library(tidyverse)
library(magrittr)
library(Seurat)
library(stringr)
library(scCustomize)

source("functions/functions_R.R")
```

[PLACEHOLDER TEXT INTRODUCTION]

## Tabula sapiens

The original Tabula Sapiens dataset consists of 58482 genes and contains a total of 483152 cells.

```{r}
tsData <- read_rds("data/raw_data/Tabula_Sapiens_all_cells.rds")

Idents(tsData) <- tsData@meta.data$cell_ontology_class

tsData@meta.data <- tsData@meta.data %>%
  rename(nFeature_RNA = nFeaturess_RNA, nCount_RNA = nCounts_RNA_UMIs)

tsMetadata <- tsData@meta.data

gc()
```

### Grouping into major cell types

```{r}
freq <- as_tibble(table(original = tsMetadata$cell_ontology_class))

tsCelltypes <- tsMetadata %>%
  distinct(cell_ontology_class, .keep_all = TRUE)%>%
  select(cell_ontology_class, compartment) %>%
  rename(original = cell_ontology_class) %>%
  left_join(freq, by = "original")

# CHECK FOR MESENCHYMAL CELL/FIBROBLAST IF THE PATTERN IS RIGHT
celltypePatterns <- list(
  # "Monocyte" = "(?i)dendritic|monocyte|macrophage|langerhans",
  "Macrophage" = "(?i)monocyte|macrophage",
  "Dendritic cell" = "(?i)dendritic|langerhans",
  # "Mesenchymal cell/fibroblast" = "(?i)limbal|  stromal|tendon|fibroblast|stellate|keratocyte|adventitial|mesenchymal",
  # "Skeletal muscle cell" = "(?i)[^hc] muscle",
  "B cell" = "(?i)(^|\\b)b\\scell|plasma[^c]",
  # "Platelet" = "(?i)platelet",
  "Basophil/mast cell" = "(?i)basophil|mast",
  # "Neutrophil" = "(?i)neutrophil",
  # "Erythrocyte" = "(?i)erythrocyte",
  # "Cardiac muscle cell" = "(?i)cardiac muscle",
  "Pericyte/smooth muscle cell" = "(?i)smooth|myometrial|pericyte",
  "T/NK cell" = "(?i)(nk)|(natural)|((\\A|\\W)t cell)",
  # "Kidney epithelial cell" = "(?i)kidney",
  # "Enterocyte" = "(?i)^(enterocyte|mature enterocyte)",
  # "Hepatocyte" = "(?i)hepatocyte", # In Seurat Object, remove the ones from heart.
  # "Pancreatic epithelial cell" = "(?i)pancreatic (ductal|acinar)",
  # "Pneumocyte" = "(?i)pneumocyte",
  # "Endothelial cell" = function(data) as.vector(data$original[data$compartment == "endothelial"])
  "Endothelial cell" = "(?i)[^(ic)][cyn]\\sendothelial cell|artery|tree|vein"
)

# CHANGE THIS PART TO SOMETHING THAT I UNDERSTAND
cluster <- lapply(celltypePatterns, function(pattern) {
  if (is.function(pattern)) {
    pattern(tsCelltypes)
  } else{
    str_subset(tsCelltypes$original, pattern)
  }
})

replaceCelltype <- tsCelltypes$original

for (name in names(cluster)) {
  replaceCelltype[replaceCelltype %in% unlist(cluster[[name]])] <- name
}

tsCelltypes$cluster7 <- factor(replaceCelltype, levels = names(cluster))

# tsCelltypes$cluster13 <- tsCelltypes$cluster17
# 
# epithelialCells <- c("Kidney epithelial cell", "Enterocyte", "Hepatocyte", "Pancreatic epithelial cell", "Pneumocyte")
# 
# tsCelltypes <- tsCelltypes %>%
#   mutate(cluster13 = ifelse(cluster13 %in% epithelialCells, "Epithelial cell", as.character(cluster13)))
# 
# tsCelltypes$cluster13 <- as.factor(tsCelltypes$cluster13)
```

Below, the metadata file of the Seurat object will be adapted to include the major cell types.

```{r}
# Rename celltypes.
tsMetadata$cell_ontology_class <- as.character(tsMetadata$cell_ontology_class)

tsMetadata$cluster7 <- tsMetadata$cell_ontology_class

for (name in names(cluster)) {
  cells <- unlist(cluster[[name]])

  tsMetadata$cluster7[tsMetadata$cluster7 %in% cells] <- name
}

tsMetadata$cluster7[!(tsMetadata$cluster7 %in% names(cluster))] <- NA

# tsMetadata$cluster13 <- tsMetadata$cluster17
# 
# tsMetadata <- tsMetadata %>%
#   mutate(cluster13 = ifelse(cluster13 %in% epithelialCells, "Epithelial cell", cluster13))

tsData@meta.data <- tsMetadata

# Remove certain organs from the metadata file.
organsToInclude <- c("Blood", "Liver", "Lymph_Node", "Fat", "Skin", "Bone_Marrow", "Heart", "Muscle", "Pancreas", "Muscle", "Thymus", "Large_Intestine", "Lung", "Small_Intestine", "Vasculature", "Kidney")

tsMetadataFilter <- tsMetadata %>%
  filter(tissue_in_publication %in% organsToInclude)

# %>%
#   filter(!(tissue_in_publication == "Heart" & cluster17 == "Hepatocyte"))
```

### Subset Tabula Sapiens into smaller Seurat objects

As processing the original Tabula Sapiens Seurat object is too memory-heavy, cells with a larger cell type frequency than 1000 will be selected at random to be put into a new Seurat object.

```{r}
nCelltypes <- "7"

newFreq <- as_tibble(table(cluster7 = tsMetadataFilter$cluster7))

cellsToSample <- newFreq$cluster7[newFreq$n > 1000]

cellsToInclude <- tsMetadataFilter %>%
  subset(cluster7 == "Dendritic cell") %>%
  rownames()

# cellsToInclude <- tsMetadataFilter %>%
#   subset(cluster7 == "Platelet") %>%
#   rownames()

for(i in 1:1) {
  sampledCells <- c()
  
  for (cell in cellsToSample) {
    rows <- tsMetadataFilter %>%
      subset(cluster7 == cell) %>%
      sample_n(1000) %>%
      rownames()
    
    sampledCells <- c(sampledCells, rows)
  }
  
  name <- paste0("cellsToInclude", i)
  
  assign(name, c(cellsToInclude, sampledCells))
  
}

for(n in 1:1) {
  listName <- paste0("cellsToInclude", n)
  objName <- paste0("tsCluster", nCelltypes, n)
  
  assign(objName, subset(tsData, cells = get(listName)))

  obj <- get(objName)
  Idents(obj) <- obj@meta.data$cluster7

  assign(objName, obj)
}
```

### Match to bulk RNA-seq data

#### Atherosclerotic plaque

The bulk RNA-seq of the atherosclerotic plaques of the AtheroExpress cohort has 19124 genes.

#### Circulating cell-free RNA

```{r}
# load("data/raw_data/cfRNA_blood_pilot.RData")
# 
# completeHGNC <- read_delim("input/hgnc_complete_set.txt")
# ncRNA <- read_delim("input/HGNC_NonCodingRNA.txt")
# 
# annotation <- read_delim(file = "input/Data_cfRNA_plasma_blood/rawCounts_combined_annotation.txt")
# 
# bulkBlood <- rawCounts_combined %>% 
#   rownames_to_column("ensembl_gene_id") %>%
#   filter(ensembl_gene_id %in% completeHGNC$ensembl_gene_id) %>%
#   left_join(select(completeHGNC, ensembl_gene_id, symbol), by = "ensembl_gene_id") %>%
#   select(ensembl_gene_id, symbol, everything()) %>%
#   filter(!(symbol %in% ncRNA$Symbol))
# 
# save(bulkBlood, file = "data/processed_data/export_files/intermediate_R_objects/pilot_blood_bulk_data_HGNC.RData")
```

```{r}
load("data/processed_data/export_files/intermediate_R_objects/AE_bulk_data_654patients_filterRNA_notfinal.RData")

# load("data/processed_data/export_files/intermediate_R_objects/pilot_blood_bulk_data_HGNC.RData")

geneKeep <- intersect(rownames(tsCluster71), bulkFilter$ensembl_gene_id) # 32397 genes.

geneMap <- bulkFilter %>%
  filter(ensembl_gene_id %in% geneKeep) %>%
  select(ensembl_gene_id, symbol)

bulkFinal <- bulkFilter %>%
  filter(ensembl_gene_id %in% geneKeep) %>%
  column_to_rownames("symbol") %>%
  select(-ensembl_gene_id)

for(n in 1:1) {
  objName <- paste0("tsCluster", nCelltypes, n)
  
  obj <- get(objName)
  obj <- subset(obj, features = geneKeep)
  assign(objName, obj)
}

rm(list = setdiff(ls(), c(paste0("tsCluster", nCelltypes, n), "bulkFinal", "nCelltypes", "geneMap")))
gc()
```

### Format files for Scaden

Currently, the features in the Seurat objects are labelled as Ensembl gene IDs. These will be changed to the corresponding gene symbols.

```{r}
# ----- Generate count files with gene symbols for Scaden
map <- setNames(geneMap$symbol, geneMap$ensembl_gene_id)

for(n in 1:1) {
  objName <- paste0("tsCluster", nCelltypes, n)
  obj <- get(objName)
  
  countsName <- paste0("tsCounts", nCelltypes, n)
  counts <- obj[["RNA"]]$counts %>%
    as.data.frame() %>%
    t()

  colnames(counts) <- map[colnames(counts)]
  
  assign(countsName, counts)
}

# ----- Generate celltype files for Scaden

for(n in 1:1) {
  objName <- paste0("tsCluster", nCelltypes, n)
  obj <- get(objName)
  
  celltypesName <- paste0("tsCellTypes", nCelltypes, n)
  celltypes <- Idents(obj) %>%
    as.data.frame() %>%
    rename(., Celltype = `.`)
  
  assign(celltypesName, celltypes)
}

table(rownames(tsCounts71) == rownames(tsCellTypes71))
```

```{r}
# write.table(bulkFinal,
#             file = "output/scaden_cfRNA/format_files/scaden_plaque_for_cfRNA_bulk_data.txt",
#             sep = "\t")

write.table(tsCellTypes71,
            file = "data/processed_data/scaden_sc_rna_ref_files/rTS_bAEplaque_c7_v1/rTS_bAEplaque_c7_v1_celltypes.txt",
            sep = "\t", row.names = FALSE)

write.table(tsCounts71,
            file = "data/processed_data/scaden_sc_rna_ref_files/rTS_bAEplaque_c7_v1/rTS_bAEplaque_c7_v1_counts.txt",
            sep = "\t", row.names = FALSE)
```

## Plots of the Seurat object

### UMAP

```{r}
tiff(paste0("output/figures/Tabula_Sapiens/TS_cluster_17_v2.tiff"), width = 12, height = 8, units = "in", res = 300)

DimPlot_scCustom(tsCluster171,
                 reduction = "umap",
                 label.box = TRUE,
                 label.size = 3,
                 repel = TRUE)

dev.off()
```

### Sankey diagram

```{r}
metadata <- tsCluster171@meta.data %>%
  select(tissue_in_publication, cluster17)

tsOrgan171 <- as.data.frame(table(metadata)) %>%
  filter(Freq >= 50)

nodes <- data.frame(
  name = c(as.character(tsOrgan171$cluster17), 
           as.character(tsOrgan171$tissue_in_publication)) %>% 
              unique()
)

tsOrgan171$IDsource <- match(tsOrgan171$cluster17, nodes$name)-1
tsOrgan171$IDtarget <- match(tsOrgan171$tissue_in_publication, nodes$name)-1

p <- networkD3::sankeyNetwork(
  Links = tsOrgan171,
  Nodes = nodes,
  Source = "IDsource",
  Target = "IDtarget",
  Value = "Freq",
  NodeID = "name",
  sinksRight = FALSE
)

htmlwidgets::saveWidget(p, file="output/figures/HTMLWidget/tsCluster171_organ_distribution.html")
```

## Session info

```{r}
devtools::session_info()
```
