---
title: "01_preprocessing_bulk_RNA"
author: "Yayuan Zhu"
date: "2024-07-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.1 Introduction

------------------------------------------------------------------------

**Goal**: Preprocess bulk RNA-seq datasets, ensuring they are ready for formatting according to the input specifications of Scaden in the next R Markdown.

**Overview workflow**:

-   For **all** the bulk RNA-seq datasets, remove genes without HGNC symbols and non-coding RNA genes.

-   If multiple bulk RNA-seq datasets need to be combined, both datasets will be filtered for lowly-expressed genes and normalized. After combination, batch effect correction will be applied.

**Note**: Users can adapt the code in chunks or lines marked with `# MODIFY HERE`. Other parts not annotated, such as file import and export, may also require changes to fit ones situation.

**Libraries**:

```{r}
library(tidyverse)

library(edgeR)
library(sva)

source("functions/functions_01_R.R")
```

# 1.2 Import files

------------------------------------------------------------------------

```{r}
# ---- HGNC symbols and non-coding RNA data ----
completeHGNC <- read.delim("data/raw_data/hgnc_complete_set.txt")
ncRNA <- read.delim("data/raw_data/HGNC_NonCodingRNA.txt")
```

```{r}
# ---- Bulk RNA-seq dataset ----

# [Athero-Express atherosclerotic plaque]
# aePlBulk1 <- read.delim("data/raw_data/AE_bulk_RNA_batch1.minRib.PC_07042023.txt") %>%
#   rename("ensembl_gene_id" = "gene")
# 
# aePlBulk2 <- read.delim("data/raw_data/AE_bulk_RNA_batch2.minRib.PC_02082023.txt") %>%
#   rownames_to_column("ensembl_gene_id")

# [Pilot ccfRNA]
# load("data/raw_data/cfRNA_blood_pilot.RData")
# pilotCcfBulk <- rawCounts_combined %>%
#   rownames_to_column("ensembl_gene_id")

# [AtheroExpress ccfRNA]
aeCcfBulk1 <- read.delim("data/raw_data/MM_MOK9065_combined_raw_counts.txt") %>%
  rename("ensembl_gene_id" = "gene")
colnames(aeCcfBulk1) <- gsub("AE", "ae", colnames(aeCcfBulk1))

aeCcfBulk2 <- read.delim("data/raw_data/cfRNA_24042024_combined_raw_counts.txt") %>%
  rename("ensembl_gene_id" = "gene")
colnames(aeCcfBulk2) <- gsub("AE", "ae", colnames(aeCcfBulk2))
colnames(aeCcfBulk2) <- gsub(".sam.counts", "", colnames(aeCcfBulk2))

# [GTEx bulk RNA-seq datasets]
# organToImport <- "skeletal_muscle"
# 
# nameDf <- paste0("gtex", str_to_title(organToImport))
# df <- readRDS(paste0("data/raw_data/GTEx_bulk_subset_", organToImport, ".rds")) %>%
#     as.data.frame() %>%
#     rownames_to_column("ensembl_gene_id") %>%
#     mutate(ensembl_gene_id = gsub("\\.\\d+", "", ensembl_gene_id)) %>%
#     select(ensembl_gene_id, everything())
# assign(nameDf, df)
# rm(df)
```

If the bulk RNA-seq file contains the final processed counts, continue to Section 1.3. If the bulk RNA-seq file(s) require(s) additional filtering, normalization, and/or integration, continue to Section 1.4.

# 1.3 Filter final bulk RNA-seq data on HGNC symbols and non-coding RNA genes

------------------------------------------------------------------------

```{r}
bulkHgncNc <- filterBulkGenes(bulkData = pilotCcfBulk, hgncData = completeHGNC, ncData = ncRNA) # MODIFY HERE

# saveRDS(bulkHgncNc, file = "data/preprocessed_data/bPilotCcf.rds")
```

# 1.4 Bulk RNA-seq data refinement and integration

------------------------------------------------------------------------

## 1.4.1 Filter genes on HGNC symbols and non-coding RNA genes

```{r}
multipleFiles <- TRUE # MODIFY HERE; one file = FALSE

bulkHgncNc1 <- filterBulkGenes(bulkData = aeCcfBulk1, hgncData = completeHGNC, ncData = ncRNA) %>% # MODIFY HERE
  column_to_rownames("ensembl_gene_id")

if(multipleFiles) {
  bulkHgncNc2 <- filterBulkGenes(bulkData = aeCcfBulk2, hgncData = completeHGNC, ncData = ncRNA) %>% # MODIFY HERE
    column_to_rownames("ensembl_gene_id")
  
  bulkHgncNcBatches <- list(bulkHgncNc1, bulkHgncNc2)
} else {
  bulkHgncNcBatches <- list(bulkHgncNc1)
}
```

```{r}
# Preparation for subsequent steps and gaining insights into the base files
for(i in seq_along(bulkHgncNcBatches)) {
  dgeName <- paste0("dge", i)
  cpmName <- paste0("cpm", i)
  
  dge <- DGEList(counts = bulkHgncNcBatches[[i]])
  cpm <- cpm(dge)
  cpmLog <- cpm(dge, log = TRUE)
    
  assign(dgeName, dge)
  assign(cpmName, cpm)

  # Distribution of expression levels across patients
  plotDensities(as.data.frame(cpmLog),
                legend = FALSE,
                main = paste0("Base CPM density plot"))
  
  # PCA plot
  pcaPlot(cpmLog, title = paste0("PCA plot of cpm bulkHgncNc", i))
}

if(multipleFiles) {
  dgeBatches <- list(dge1, dge2)
  cpmBatches <- list(cpm1, cpm2)
} else {
  dgeBatches <- list(dge1)
  cpmBatches <- list(cpm1)
}

rm(dge, cpm, dgeName, cpmName, cpmLog)
```

## 1.4.2 Filter out lowly-expressed genes

| Bulk RNA-seq data batches    | Gene expressed | Gene expressed in `x` patients (\~10%)            | Patient expressed `x` genes (histogram of expressed genes) |
|-----------------|-----------------|------------------|---------------------|
| AtheroExpress plaque batches | `cpm > 1`      | 50 (batch 1: 654 patients, batch 2: 471 patients) | 15300                                                      |
| AtheroExpress ccfRNA batches | `cpm > 1`      | 15 (batch 1: 150 patients, batch 2: 248 patients) | 8000                                                       |

**Note**: for consistency between the batches, identical thresholds will be applied to both batches. The thresholds selected are balanced on the characteristics of the two batches.

```{r}
# Determine minimum number of genes one patient needs to express
lapply(cpmBatches, function(c) {
  ggplot(as.data.frame(colSums(c > 1)), aes(x = colSums(c > 1))) +
    geom_histogram() +
    geom_vline(xintercept = 8000) +
    labs(x = "# of genes expressed",
         y = "Frequency in patients")
})
```

```{r}
for(i in seq_along(dgeBatches)) {
  dgeHighExprName <- paste0("dgeHighExpr", i)
  
  # Genes need to be expressed (cpm > 1) in at least x patients
  toKeepGenes <- rowSums(cpmBatches[[i]] > 1) >= 15 # MODIFY HERE
  # Patients need to have at least x genes expressed
  toKeepPatients <- colSums(cpmBatches[[i]] > 1) >= 8000 # MODIFY HERE

  dgeHighExpr <- dgeBatches[[i]][toKeepGenes, toKeepPatients, keep.lib.sizes = FALSE]
  cpmHighExpr <- cpm(dgeHighExpr, log = TRUE)

  assign(dgeHighExprName, dgeHighExpr)

  # Distribution of expression levels across patients
  plotDensities(as.data.frame(cpmHighExpr),
                legend = FALSE, main = "Filtered lowly expressed genes CPM density plot")
  
  # PCA plot
  pcaPlot(cpmHighExpr, title = paste0("PCA plot of cpm bulkHighExpr", i))
}

if(multipleFiles) {
  dgeHighExprBatches <- list(dgeHighExpr1, dgeHighExpr2)
} else {
  dgeHighExprBatches <- list(dgeHighExpr1)
}

rm(dgeHighExpr, dgeHighExprName, cpmHighExpr)
```

## 1.4.3 Normalization

```{r}
for(i in seq_along(dgeHighExprBatches)) {
  normName <- paste0("dgeNorm", i)
  
  dgeNorm <- calcNormFactors(dgeHighExprBatches[[i]], method = "TMM")
  cpmNorm <- cpm(dgeNorm, log = TRUE)
  
  assign(normName, dgeNorm)
  
  # Distribution of expression levels across patients
  plotDensities(as.data.frame(cpmNorm),
                legend = FALSE, main = "Normalization CPM density plot")
  
  # PCA plot
  pcaPlot(cpmNorm, title = paste0("PCA plot of cpm normalized", i))
}

if(multipleFiles) {
  dgeNormBatches <- list(dgeNorm1, dgeNorm2)
} else {
  dgeNormBatches <- list(dgeNorm1)
}

rm(dgeNorm, cpmNorm)
```

## 1.4.4 Combining normalized counts of batches and correcting for batch effect

```{r}
bulkNorm <- lapply(dgeNormBatches, function(dge) {
  cpm(dge)
})

bulkNorm1 <- as.data.frame(bulkNorm[[1]]) %>%
  rownames_to_column("ensembl_gene_id") %>%
  left_join(select(completeHGNC, ensembl_gene_id, symbol), by = "ensembl_gene_id") %>%
  select(ensembl_gene_id, symbol, everything())

if(multipleFiles) {
  bulkNorm2 <- as.data.frame(bulkNorm[[2]]) %>%
  rownames_to_column("ensembl_gene_id") %>%
  left_join(select(completeHGNC, ensembl_gene_id, symbol), by = "ensembl_gene_id") %>%
  select(ensembl_gene_id, symbol, everything())
}
```

```{r}
combineBulkNorm <- inner_join(bulkNorm1, bulkNorm2, by = c("ensembl_gene_id", "symbol"), suffix = c("_batch1", "_batch 2"))
  
metadataCombine <- bind_rows(data.frame(Patient = colnames(bulkNorm1)[-c(1, 2)], batch = as.factor("1")),
                             data.frame(Patient = colnames(bulkNorm2)[-c(1, 2)], batch = as.factor("2")))

table(duplicated(metadataCombine$Patient))
```

```{r}
combineBulkNormPCA <- combineBulkNorm %>%
  column_to_rownames("ensembl_gene_id") %>%
  select(-symbol)

pca <- prcomp(t(combineBulkNormPCA), scale = TRUE)
pcaData <- as.data.frame(pca$x)
pcaData$batch <- metadataCombine$batch

# Initial evaluation of batch effect
ggplot(pcaData, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = batch)) +
  labs(title = "PCA of uncorrected bulk RNA-seq Data",
       x = paste0("PC1: ", round(summary(pca)$importance[2,1] * 100, 2), "% Variance"),
       y = paste0("PC2: ", round(summary(pca)$importance[2,2] * 100, 2), "% Variance")) +
  theme_minimal()

# Correct for batch effect
batchCorrectedBulk <- ComBat_seq(counts = as.matrix(combineBulkNormPCA), batch = metadataCombine$batch)

pcaCorrected <- prcomp(t(batchCorrectedBulk), scale = TRUE)
pcaDataCorrected <- as.data.frame(pcaCorrected$x)
pcaDataCorrected$batch <- metadataCombine$batch

ggplot(pcaDataCorrected, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = batch)) +
  labs(title = "PCA of batch-corrected bulk RNA-seq Data",
       x = paste0("PC1: ", round(summary(pcaCorrected)$importance[2,1] * 100, 2), "% Variance"),
       y = paste0("PC2: ", round(summary(pcaCorrected)$importance[2,2] * 100, 2), "% Variance")) +
  theme_minimal()
```

```{r}
bulkCombineCorrected <- as.data.frame(batchCorrectedBulk) %>%
  rownames_to_column("ensembl_gene_id") %>%
  left_join(select(completeHGNC, ensembl_gene_id, symbol), by = "ensembl_gene_id") %>%
  select(ensembl_gene_id, symbol, everything())
```

## 1.4.5 Export preprocessed bulk RNA-seq file(s)

```{r}
saveRDS(bulkNorm1, "data/preprocessed_data/bAEccfRNA1_v2.rds")

if(multipleFiles) {
  saveRDS(bulkNorm2, "data/preprocessed_data/bAEccfRNA2_v2.rds")
}

if(exists("bulkCombineCorrected")) {
   # Corrected for batches
} else if(exists("combineBulkNorm")) {
  saveRDS(combineBulkNorm, "data/preprocessed_data/bAEccfRNACombine_v2.rds") # Uncorrected for batches
}
```
