---
title: "Analysis_prediction_Scaden"
author: "Yayuan Zhu"
date: "2024-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup}
library(tidyverse)
library(magrittr)
library(reshape2)
library(hrbrthemes)
library(viridis)
library(ggthemes)
extrafont::loadfonts()

source("functions/functions_R.R")
```

[PLACEHOLDER TEXT INTRODUCTION]

## Obtaining average proportions and corresponding graphs

```{r}
pattern <- "rTS_bAEplaque_c13_v1"
nCelltypes <- "13"
path <- paste0("output/scaden_predictions/", pattern, "/", pattern, "_n")

# pattern <- "rAEplaque_bAEplaque_c7"
# path <- paste0("output/scaden_predictions/", pattern, "/predictions_", pattern, "_s1000_n")

nSimulations <- seq_len(10)

listLongProportions <- getLongProportions(path, nSimulations)
avgProportions <- getAvgProportions(listLongProportions, nSimulations)
```

### Graphs of all and average predicted cell type proportions

```{r}
#All predictions
titlePlotAll = paste0("Cellular deconvolution predictions: ", pattern)

pdf(paste0("output/pdf/", pattern, "_all_graphs.pdf"))

for(n in nSimulations) {
  longFile <- paste0("longPred", n)

  p <- ggplot(listLongProportions[[longFile]], aes(x = Celltype, y = Proportion)) +
    geom_boxplot() +
    labs(title = titlePlotAll,
         subtitle = paste("Celltypes:", nCelltypes,
                          "\nRun:", n)) +
    theme(axis.text.x = element_text(angle=60, vjust = 1, hjust = 1, size = 8)) +
    # scale_y_continuous(name = "Average cell type proportion",
    #                    limits = c(0, 0.8))
    scale_y_continuous(name = "Relative average cell type proportion")

  print(p)
}

dev.off()

#Average prediction
titlePlotAvg = paste0("Cellular deconvolution average prediction: ", pattern)

tiff(paste0("output/plots/average_predictions_plots/", pattern, "_avg_prediction.tiff"),
     width = 12,
     height = 12,
     units = "in",
     res = 300)

pAvg <- ggplot(avgProportions, aes(x = reorder(Celltype, `Average proportion`), y = (`Average proportion`))) +
  geom_boxplot(fill = "lightgrey") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     breaks = seq(0, 1, by = 0.1)) +
  labs(title = titlePlotAvg,
       subtitle = paste("Cell types:", nCelltypes,
                        "\nTotal runs:", length(nSimulations))) +
  ylab("Relate average proportion") +
  xlab("Cell type") +
  theme_ipsum_rc(grid="Y") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x = element_text(size=13, hjust=0.5, margin=margin(t=15)),
        axis.title.y = element_text(size=13, hjust=0.5, margin=margin(r=15)),
        panel.grid.minor.y = element_line(color="gray", size=0.5),
        legend.title = element_text(size=13),
        legend.text = element_text(size = 11))

print(pAvg)

dev.off()
```

### Comparing different groups

Below, predictions of different versions of cellular predictions can be compared to each other. The versions need to have the same source of the reference, same bulk RNA-seq data, and the same amount of cells and cell types used.

```{r}
avgPropCluster14 <- avgProportions1 %>%
  left_join(avgProportions2, by = c("Patient", "Celltype")) %>%
  left_join(avgProportions3, by = c("Patient", "Celltype")) %>%
  left_join(avgProportions4, by = c("Patient", "Celltype")) %>%
  rename(`1` = `Average proportion.x`,
         `2` = `Average proportion.y`,
         `3` = `Average proportion.x.x`,
         `4` = `Average proportion.y.y`)

avgPropCluster14_long <- avgPropCluster14 %>%
  pivot_longer(cols = c(`1`, `2`, `3`, `4`),
               names_to = "Group",
               values_to = "Average proportion") %>%
  mutate(Celltype = as.factor(Celltype),
         Group = as.factor(Group),
         Patient = as.factor(Patient))

tiff("output/figures/TS_comparison_randomization_c14_v4_violin.tiff", width = 12, height = 8, units = "in", res = 300)

ggplot(avgPropCluster14_long, aes(fill=Group, x=Celltype, y=`Average proportion`)) + 
  geom_violin(position=position_dodge(width=0.8), alpha=0.5) +
  scale_fill_viridis(discrete=TRUE, name="Randomization") +
  scale_y_percent() +
  ylab("Relative average proportion") +
  xlab("Celltype") +
  labs(title = "Average predicted cellular proportions: atherosclerotic plaque",
       subtitle = "Reference: Tabula Sapiens") +
  theme_ipsum_rc(grid="Y") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x = element_text(size=13, hjust=0.5, margin=margin(t=15)),
        axis.title.y = element_text(size=13, hjust=0.5, margin=margin(r=15)),
        panel.grid.minor.y = element_line(color="gray", size=0.5),
        legend.title = element_text(size=13),
        legend.text = element_text(size = 11))

dev.off()

tiff("output/figures/TS_comparison_randomization_c14_v4_boxplot.tiff", width = 12, height = 8, units = "in", res = 300)

ggplot(avgPropCluster14_long, aes(fill=Group, x=Celltype, y=`Average proportion`)) + 
  geom_boxplot(position=position_dodge(width=0.8), width=0.5, alpha=0.5, outlier.shape = NA) +
  scale_fill_viridis(discrete=TRUE, name="Randomization") +
  scale_y_percent() +
  labs(title = "Average predicted cellular proportions: atherosclerotic plaque",
       subtitle = "Reference: Tabula Sapiens") +
  ylab("Relative average proportion") +
  xlab("Celltype") +
  theme_ipsum_rc(grid="Y") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x = element_text(size=13, hjust=0.5, margin=margin(t=15)),
        axis.title.y = element_text(size=13, hjust=0.5, margin=margin(r=15)),
        panel.grid.minor.y = element_line(color="gray", size=0.5),
        legend.title = element_text(size=13),
        legend.text = element_text(size = 11))

dev.off()
```

#### Statistics

```{r}
# Test for normality
shapiroResults <- shapiroTestCustom(avgPropCluster13_long) # All celltypes no normality.

celltypesQQPlot(avgPropCluster13_long)

# Test for equality of variances
leveneResults <- leveneTestCustom(avgPropCluster13_long) # Homogenous variances: basophil/mast cell, mesenchymal cell/fibroblast, and pericyte/smooth muscle cell.

# Kruskal-Wallis test to see if there is a difference between each group, within each celltype.
kruskalResults <- kruskalTestCustom(avgPropCluster13_long)

# Post-hoc test: Dunn's test
dunnResults <- dunnTestCustom(data = avgPropCluster13_long) # 4/42 comparisons non-significant.
```

## Patient data

To the AtheroExpress patient data, the columns containing the characteristics of interest will be altered in order to make it more readable.

```{r}
patientData <- haven::read_sav("input/AEdata03082023.sav")

patientData$STUDY_NUMBER <- gsub("^", "ae", patientData$STUDY_NUMBER)

# Use the code below for finding the meaning behind the numbers in the column.
# labelled::val_labels(patientData$Gender)

patientData["Symptoms.2g"] <- case_when(patientData$Symptoms.4g == 0 | patientData$Symptoms.4g == 1~"mild",
                                        patientData$Symptoms.4g == 2 | patientData$Symptoms.4g == 3~"severe")

patientData["Symptoms.4g"] <- case_when(patientData$Symptoms.4g == 0~"asymptomatic",
                                        patientData$Symptoms.4g == 1~"ocular",
                                        patientData$Symptoms.4g == 2~"TIA",
                                        patientData$Symptoms.4g == 3~"stroke")

patientData["Gender"] <- case_when(patientData$Gender == 0~"female",
                                   patientData$Gender ==1~"male")

# save(patientData, file = "output/export_files/intermediate_R_objects/AE_patientdata_with_symptoms2g_symptoms4g_gender.rds")
```

```{r}
load("output/export_files/intermediate_R_objects/AE_patientdata_with_symptoms2g_symptoms4g_gender.rds")
```

### Match patient data to cellular deconvolution results

**Make functions where you can more easily add the symptoms to the corresponding patient!!!!**

```{r}
allPatients <- patientData %>%
  filter(STUDY_NUMBER %in% avgProportions$Patient) %>%
  select(STUDY_NUMBER, Symptoms.4g, Symptoms.2g, Gender) %>%
  rename(Patient = STUDY_NUMBER) %>%
  na.omit()

avgProp2g <- matchSymptomGroup(allPatients, avgProportions, "Symptoms.2g", c("mild", "severe")) %>%
  bind_rows()

avgProp4g <- matchSymptomGroup(allPatients, avgProportions, "Symptoms.4g", c("asymptomatic", "ocular", "TIA", "stroke")) %>%
  bind_rows()

avgPropGender <- matchSymptomGroup(allPatients, avgProportions, "Gender", c("female", "male")) %>%
  bind_rows()

avgProp4g$Symptoms.4g <- factor(avgProp4g$Symptoms.4g, levels = c("asymptomatic", "ocular", "TIA", "stroke"))

avgPropSymp <- avgProp2g %>%
  left_join(avgProp4g, by = c("Patient", "Celltype", "Average proportion")) %>%
  left_join(avgPropGender, by = c("Patient", "Celltype", "Average proportion"))
```

### Graph of average cellular deconvolution prediction based on patient characteristic

#### Two characteristics - scale

```{r}
typeDeconv = "cfRNA"
patientCharacteristic = "gender"
# colPatientChar = Gender

titlePlotAvg = "Cellular deconvolution average prediction: cfRNA"
subtitlePlotAvg = "Male versus female"

tiff(paste0("output/figures/", typeDeconv, "_",patientCharacteristic,".tiff"),
     width = 720,
     height = 720,
     units = "px")

pAvg2g <- ggplot(avgPropSymp, aes(x = Celltype, y = `Average proportion`, fill = Gender)) +
  geom_boxplot() +
  labs(title = paste(typeDeconv, "predictions"),
       subtitle = subtitlePlotAvg,
       fill = "Group") +
  scale_fill_manual(values = c("#8ec3de", "#f6a582")) +
  scale_color_manual(values = c("black", "black")) +
  scale_y_continuous(name = "Relative average proportion",
                     limits = c(0, 0.8)) +
  ownMinimalTheme()

print(pAvg2g)

dev.off()
```

#### Four characteristics - scale

```{r}
tiff(paste0("output/figures/", typeDeconv, "_sympt4g.tiff"),
     width = 720,
     height = 720,
     units = "px")

pAvg4g <- ggplot(avgPropSymp, aes(x = Celltype, y = `Average proportion`, fill = Symptoms.4g)) +
  geom_boxplot() +
  labs(title = paste(typeDeconv, "predictions"),
       subtitle = "Asymptomatic versus ocular versus TIA versus stroke",
       fill = "Group") +
  scale_fill_manual(values = c("#1065ab", "#8ec3de", "#f6a582", "#b31529")) +
  scale_color_manual(values = c("black", "black", "black","black")) +
  scale_y_continuous(name = "Relative average proportion",
                           limits = c(0, 0.8)) +
  ownMinimalTheme()

print(pAvg4g)

dev.off()
```

## Correlation of predictions

### Bulk RNA-seq: atherosclerotic plaque

```{r}
listLongProportionsPlaque <- getLongProportions(basePathPrediction = "output/scaden_predictions/rAEplaque_bAEplaque_c7/predictions_rAEplaque_bAEplaque_c7_s1000_n", nSimulations = seq_len(10))
avgProportionsPlaque <- getAvgProportions(nameList = listLongProportionsPlaque,
                                          nSimulations = seq_len(10))
```

#### In-house versus Tabula Sapiens reference

##### Combined epithelial cells

```{r}
combineAvgPlaque <- avgProportionsPlaque %>%
  mutate(Celltype = ifelse(Celltype %in% c("Macrophage", "Dendritic cell"), "Monocyte", Celltype)) %>%
  group_by(Patient, Celltype) %>%
  summarise(`Average proportion` = mean(`Average proportion`), .groups = 'drop')

filterAvgCfRNA <- avgProportions %>%
  subset(Celltype %in% c("B cell", "Basophil/mast cell", "Endothelial cell", "Monocyte", "Pericyte/smooth muscle cell", "T/NK cell"))

celltypeMapping <- data.frame(plaqueCelltype = c("Endothelial cell", "Mast cell", "Memory B cell", "Monocyte", "Smooth muscle cell", "T/NK cell"),
  cfRNACelltype = c("Endothelial cell", "Basophil/mast cell", "B cell", "Monocyte", "Pericyte/smooth muscle cell", "T/NK cell")
)

test <- combineAvgPlaque %>%
  left_join(celltypeMapping, by = c("Celltype" = "plaqueCelltype"))

test2 <- filterAvgCfRNA %>%
  left_join(celltypeMapping, by = c("Celltype" = "cfRNACelltype"))

merge <- test %>%
  select(Patient, Celltype = cfRNACelltype, `Average proportion plaque` = `Average proportion`) %>%
  inner_join(test2 %>%
               select(Patient, Celltype, `Average proportion Tabula Sapiens` = `Average proportion`),
             by = c("Patient", "Celltype"))

tiff("output/figures/cor_plaque_cfRNA_c17_v1.tiff", width = 12, height = 8, units = "in", res = 300)

ggplot(merge, aes(x = `Average proportion plaque`, y = `Average proportion Tabula Sapiens`)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average predicted cellular proportions: atherosclerotic plaque ",
       subtitle = "Reference: atherosclerotic plaque versus Tabula Sapiens\nRandomization: 3") +
  theme_ipsum_rc(grid="Y") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x = element_text(size=13, hjust=0.5, margin=margin(t=15)),
        axis.title.y = element_text(size=13, hjust=0.5, margin=margin(r=15)),
        panel.grid.minor.y = element_line(color="gray", size=0.5),
        legend.title = element_text(size=13),
        legend.text = element_text(size = 11))

dev.off()
```

##### One Tabula Sapiens cell type versus all AE plaque cell types

```{r}
listCellTypes <- unique(avgProportions$Celltype)

combine <- avgProportionsPlaque %>%
  rename(`Average proportion Plaque` = `Average proportion`)

for(cellType in listCellTypes) {
  avgPropName <- paste("Average proportion", cellType)
  
  col <- avgProportions %>%
    filter(Celltype == cellType) %>%
    rename(!!avgPropName := `Average proportion`) %>%
    select(-Celltype)
  
  combine <- combine %>%
    left_join(col, by = "Patient")
}

plotNamesCellTypes <- listCellTypes %>%
  str_replace_all(c(" " = "_",
                    "/" = "_"))

combineRename <- combine

colnames(combineRename) <- colnames(combineRename) %>%
  str_replace_all(c(" " = "_",
                    "/" = "_"))
```

```{r}
for(i in (4:ncol(combineRename))) {
  colName <- colnames(combineRename)[i]
  
  tiff(paste0("output/plots/correlation_plots/", pattern, "/cor_", plotNamesCellTypes[i-3], "_", pattern, ".tiff"),
       width = 12,
       height = 8,
       units ="in",
       res = 300)
  
  p <- ggplot(combineRename, aes(x = .data[[colName]], y = Average_proportion_Plaque)) +
    geom_point() +
    geom_smooth(method = lm) +
    facet_wrap(~ Celltype,
               nrow = 3,
               scales = "free") +
    labs(title = paste("Correlation average predicted cellular proportions", pattern),
           subtitle = paste("Comparison: AE plaque versus Tabula Sapiens", plotNamesCellTypes[i-3])) +
    theme_ipsum_rc(grid="Y") +
    theme(axis.text.x = element_text(angle=45, hjust=1),
            axis.title.x = element_text(size=13, hjust=0.5, margin=margin(t=15)),
            axis.title.y = element_text(size=13, hjust=0.5, margin=margin(r=15)),
            panel.grid.minor.y = element_line(color="gray", size=0.5),
            legend.title = element_text(size=13),
            legend.text = element_text(size = 11))
  
  print(p)
  
  dev.off()
}
```

```{r}
corAEplaqueTS <- combineRename %>%
  group_by(Celltype) %>%
  summarise(across(starts_with("Average_proportion_"), 
                   ~ cor(.x, Average_proportion_Plaque), .names = "cor_{col}")) %>%
  column_to_rownames("Celltype") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("TS_cell_type") %>%
  pivot_longer(cols = -TS_cell_type, names_to = "AE_plaque_cell_type", values_to = "Correlation") %>%
  mutate(TS_cell_type = gsub("cor_Average_proportion_", "", TS_cell_type))

corLevels <- unique(corAEplaqueTS$TS_cell_type)
corLevels <- corLevels[corLevels != "Plaque"]

orderLevels <- c(rev(corLevels), "", "Plaque")

dummy <- data.frame(TS_cell_type = rep("", 7),
                    AE_plaque_cell_type = unique(corAEplaqueTS$AE_plaque_cell_type),
                    Correlation = NA)

corAEplaqueTS <- rbind(corAEplaqueTS[1:7,], dummy, corAEplaqueTS[8:nrow(corAEplaqueTS),])

corAEplaqueTS$TS_cell_type <- factor(corAEplaqueTS$TS_cell_type, 
                                     levels = orderLevels)

tiff(paste0("output/plots/correlation_plots/", pattern, "_heatmap.tiff"),
     width = 12,
     height = 12,
     units = "in",
     res = 300)

ggplot(corAEplaqueTS, aes(x = AE_plaque_cell_type, y = TS_cell_type)) +
  geom_tile(aes(fill = Correlation),
            color = "white") +
  scale_fill_distiller(palette = "RdBu",
                       limit = c(-1, 1),
                       na.value="white",
                       direction = 1) +
  labs(title = "Correlation of average cellular proportions of AtheroExpress bulk RNA-seq data\nbetween Tabula Sapiens and AtheroExpress plaque references",
       subtitle = pattern,
       x = "AtheroExpress plaque reference",
       y = "Tabula Sapiens reference") +
  theme_ipsum_rc(grid = FALSE) +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x = element_text(size=13, hjust=0.5, margin=margin(t=10)),
        axis.title.y = element_text(size=13, hjust=0.5, margin=margin(r=15)),
        legend.title = element_text(size=13),
        legend.text = element_text(size = 11))

dev.off()
```

### Plaque mine vs Gemma

```{r}
filterGemmaPropSymp <- filter(gemmaPropSymp, Patient %in% avgPropSymp$Patient)

#Mild
own <- filter(avgPropSymp, Symptoms.2g == "mild")
gemma <- filter(filterGemmaPropSymp, Symptoms.2g == "mild")
compare <- own[1:2]
compare["ownProp"] <- own$`Average proportion`
compare["gemmaProp"] <- gemma$`Average proportion`

tiff("output/figures/cor_plaque_mild.tiff",
     width = 720,
     height = 720,
     units = "px")

ggplot(compare, aes(x = ownProp, y = gemmaProp)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average cellular proportions of Sympt2g_mild")

dev.off()

#Severe
own <- filter(avgPropSymp, Symptoms.2g == "severe")
gemma <- filter(filterGemmaPropSymp, Symptoms.2g == "severe")
compare <- own[1:2]
compare["ownProp"] <- own$`Average proportion`
compare["gemmaProp"] <- gemma$`Average proportion`

tiff("output/figures/cor_plaque_severe.tiff",
     width = 720,
     height = 720,
     units = "px")

ggplot(compare, aes(x = ownProp, y = gemmaProp)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average cellular proportions of Sympt2g_severe")

dev.off()

#Asymptomatic
own <- filter(avgPropSymp, Symptoms.4g == "asymptomatic")
gemma <- filter(filterGemmaPropSymp, Symptoms.4g == "asymptomatic")
compare <- own[1:2]
compare["ownProp"] <- own$`Average proportion`
compare["gemmaProp"] <- gemma$`Average proportion`

tiff("output/figures/cor_plaque_asymptomatic.tiff",
     width = 720,
     height = 720,
     units = "px")

ggplot(compare, aes(x = ownProp, y = gemmaProp)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average cellular proportions of Sympt4g_asymptomatic")

dev.off()

#Ocular
own <- filter(avgPropSymp, Symptoms.4g == "ocular")
gemma <- filter(filterGemmaPropSymp, Symptoms.4g == "ocular")
compare <- own[1:2]
compare["ownProp"] <- own$`Average proportion`
compare["gemmaProp"] <- gemma$`Average proportion`

tiff("output/figures/cor_plaque_ocular.tiff",
     width = 720,
     height = 720,
     units = "px")

ggplot(compare, aes(x = ownProp, y = gemmaProp)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average cellular proportions of Sympt4g_ocular")

dev.off()

#TIA
own <- filter(avgPropSymp, Symptoms.4g == "TIA")
gemma <- filter(filterGemmaPropSymp, Symptoms.4g == "TIA")
compare <- own[1:2]
compare["ownProp"] <- own$`Average proportion`
compare["gemmaProp"] <- gemma$`Average proportion`

test <- filter(own, ! (own$Patient %in% gemma$Patient))

tiff("output/figures/cor_plaque_TIA.tiff",
     width = 720,
     height = 720,
     units = "px")

ggplot(compare, aes(x = ownProp, y = gemmaProp)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average cellular proportions of Sympt4g_TIA")

dev.off()

#Stroke
own <- filter(avgPropSymp, Symptoms.4g == "stroke")
gemma <- filter(filterGemmaPropSymp, Symptoms.4g == "stroke")
compare <- own[1:2]
compare["ownProp"] <- own$`Average proportion`
compare["gemmaProp"] <- gemma$`Average proportion`

tiff("output/figures/cor_plaque_stroke.tiff",
     width = 720,
     height = 720,
     units = "px")

ggplot(compare, aes(x = ownProp, y = gemmaProp)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Celltype, nrow = 3, scales = "free") +
  labs(title = "Correlation average cellular proportions of Sympt4g_stroke")

dev.off()
```

## Atherosclerotic plaque

### Own predictions

...

### Gemma predictions

```{r}
#17 patients here are not present in own predictions.
gemmaProp <- subset(read_csv("input/Gemmaprops8clustersAE.csv"), Algorithm == "Scaden")

gemmaFinalProp <- gemmaProp %>%
  mutate(`Endothelial cell` = rowMeans(select(., `Endothelial Cells I`, `Endothelial Cells II`))) %>%
  select(-`Endothelial Cells I`, -`Endothelial Cells II`, -Algorithm) %>%
  select(1, order(colnames(.))) %>%
  pivot_longer(
    cols = -Patient,
    names_to = "Celltype",
    values_to = "Average proportion"
  )

tiff("output/figures/Gemma_plaque_all.tiff",
     width = 720,
     height = 720,
     units = "px")

pGemma <- ggplot(gemmaFinalProp, aes(x = Celltype, y = `Average proportion`)) +
  geom_boxplot(fill = "#98CAE1") +
  labs(title = "Gemma plaque predictions",
       subtitle = "All patients") +
  scale_y_continuous(name = "Average cell type proportion",
                     limits = c(0, 0.8)) +
  ownMinimalTheme()

print(pGemma)

dev.off()
```

### Different patient characteristics

```{r}
gemmaPatients <- patientData %>%
  filter(STUDY_NUMBER %in% gemmaFinalProp$Patient) %>%
  select(STUDY_NUMBER, Symptoms.4g, Symptoms.2g) %>%
  rename(Patient = STUDY_NUMBER) %>%
  na.omit()

gemmaProp2g <- matchSymptomGroup(gemmaPatients, gemmaFinalProp, "Symptoms.2g", c("mild", "severe")) %>%
  bind_rows()

gemmaProp4g <- matchSymptomGroup(gemmaPatients, gemmaFinalProp, "Symptoms.4g", c("asymptomatic", "ocular", "TIA", "stroke")) %>%
  bind_rows()

gemmaProp4g$Symptoms.4g <- factor(gemmaProp4g$Symptoms.4g, levels = c("asymptomatic", "ocular", "TIA", "stroke"))

gemmaPropSymp <- gemmaProp2g %>%
  left_join(gemmaProp4g, by = c("Patient", "Celltype", "Average proportion"))
```

#### Mild versus severe

```{r}
tiff("output/figures/gemma_plaque_sympt2g.tiff",
     width = 720,
     height = 720,
     units = "px")

pGemma2g <- ggplot(gemmaPropSymp, aes(x = Celltype, y = `Average proportion`, fill = Symptoms.2g)) +
  geom_boxplot() +
  labs(title = "Gemma plaque predictions",
       subtitle = "Mild versus severe symptoms",
       fill = "Group") +
  scale_fill_manual(values = c("#8ec3de", "#f6a582")) +
  scale_color_manual(values = c("black", "black")) +
  scale_y_continuous(name = "Average cell type proportion",
                     limits = c(0, 0.8)) +
  ownMinimalTheme()

print(pGemma2g)

dev.off()
```

#### Asymptomatic versus ocular versus TIA versus stroke

```{r}
tiff("output/figures/gemma_plaque_sympt4g.tiff",
     width = 720,
     height = 720,
     units = "px")

pGemma4g <- ggplot(gemmaPropSymp, aes(x = Celltype, y = `Average proportion`, fill = Symptoms.4g)) +
  geom_boxplot() +
  labs(title = "Gemma plaque predictions",
       subtitle = "Asymptomatic versus ocular versus TIA versus stroke",
       fill = "Group") +
  scale_fill_manual(values = c("#1065ab", "#8ec3de", "#f6a582", "#b31529")) +
  scale_color_manual(values = c("black", "black", "black","black")) +
  scale_y_continuous(name = "Average cell type proportion",
                           limits = c(0, 0.8)) +
  ownMinimalTheme()

print(pGemma4g)

dev.off()
```

#### Statistical test

```{r}
#Data is skewed to the left.
ggplot(gemmaPropSymp, aes(x = Symptoms.2g, y = `Average proportion`, fill = Symptoms.2g)) +
  geom_boxplot() +
  labs(x = "Group", 
       y = "Cellular Proportions")

ggplot(gemmaPropSymp, aes(x = `Average proportion`, fill = Symptoms.2g)) +
  geom_histogram(binwidth = 0.1, 
                 position = "identity", 
                 alpha = 0.7) +
  facet_wrap(~Symptoms.2g) +
  labs(x = "Cellular Proportions", 
       y = "Frequency")

ggpubr::ggqqplot(gemmaPropSymp, x = "Average proportion") +
  labs(x = "Theoretical Quantiles", 
       y = "Sample Quantiles")

# Mild vs severe
#Test for normality: no normal distribution.
resShapiroG <- gemmaPropSymp %>%
  group_by(Symptoms.2g) %>%
  summarise(p_value = shapiro.test(`Average proportion`)$p.value)

print(resShapiroG)

#Test for equality of variances: variances are equal.
resLeveneG <- var.test(`Average proportion` ~ Symptoms.2g, 
                      data = gemmaPropSymp)

print(resLeveneG)

#Test for differences between groups: no differences with p-value = 0.1683.
resWilcoxG <- wilcox.test(`Average proportion` ~ Symptoms.2g, 
                       data = gemmaPropSymp)
print(resWilcoxG)
```

## Circulating RNA

### Tabula Sapiens

#### Bulk RNA dataset - atherosclerotic plaque

## Statistical test

### Mild versus severe symptoms

```{r}
#Test for normality: no normal distribution.
resShapiro <- avgPropSymp %>%
  group_by(Symptoms.2g) %>%
  summarise(p_value = shapiro.test(`Average proportion`)$p.value)

print(resShapiro)

#Test for equality of variances: variances are equal.
resLevene <- var.test(`Average proportion` ~ Symptoms.2g, 
                      data = avgPropSymp)

print(resLevene)

#Test for differences between groups: no differences with p-value = 0.6846.
resWilcox <- wilcox.test(`Average proportion` ~ Symptoms.2g, 
                       data = avgPropSymp)
print(resWilcox)
```

## Session info

```{r}
devtools::session_info()
```
