---
title: "05_clinical_correlations"
author: "Yayuan Zhu"
date: "2024-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 5.1 Introduction

------------------------------------------------------------------------

...

# 5.2 Libraries

------------------------------------------------------------------------

```{r}
library(tidyverse)
library(broom)
library(gtools) # Adds significance stars to tidy()
library(rlang)

library(rstatix)
library(survival) # coxzph
library(ranger)

library(survminer) # ggcoxzph
library(patchwork)
# extrafont::font_import(pattern = "Helvetica")
extrafont::loadfonts(quiet = TRUE)

source("functions/functions_05_R.R")
source("functions/functions_global_R.R")
```

# 5.3 (Average) predicted cellular proportions and corresponding patient data

## 5.3.1 Patient data

### 5.3.1.1 AtheroExpress cohort

```{r}
patientData <- haven::read_sav("data/metadata/AE_patient_data_20230803.sav") %>%
  rename(patient = STUDY_NUMBER)

patientData$patient <- gsub("^", "ae", patientData$patient)

# Add the Symptoms.2g column
patientData$Symptoms.2g <- case_when(patientData$Symptoms.4g == 0 | patientData$Symptoms.4g == 1~0,
                                     patientData$Symptoms.4g == 2 | patientData$Symptoms.4g == 3~1)

labelled::val_labels(patientData$Symptoms.2g) <- c(mild = 0, severe = 1)

# Add the thrombus.bin column
patientData$thrombus.bin <- case_when(patientData$thrombus == 0 | patientData$thrombus == 1~0,
                                     patientData$thrombus == 2 | patientData$thrombus == 3~1,
                                     patientData$thrombus == -888~-888)

labelled::val_labels(patientData$thrombus.bin) <- c(`no/minor` = 0, `moderate/heavy` = 1, `not judgeable` = -888)

# Alter order of the factors
patientData$smc_macrophages_ratio <- factor(patientData$smc_macrophages_ratio, levels = c(1, 3, 2))
```

## 5.3.2 (Average) predicted cellular proportions

```{r}
proportionsAsFractions <- TRUE # MODIFY HERE; proportions as fractions are the default
zTransform <- TRUE # MODIFY HERE; set to FALSE for Section 5.5.1

pattern <- "rTS_bAEccfRNACombine_c17_v1" # MODIFY HERE

pred <- readRDS(paste0("output/export_files/rds_files/scaden_wide_avg_prop_", pattern, ".rds")) %>% # MODIFY HERE; as wide format
  rename_with(.fn = ~ str_replace_all(., c(" " = "_", "/" = "_")),
              .cols = -patient)

# Optional modification
predModWide <- pred

  # < Convert proportions to percentages >
if(!proportionsAsFractions) {
  predModWide <- predModWide %>%
    mutate(across(-patient, ~ . * 100))
}

  # < Transform predicted cellular proportions to Z scores >
if(zTransform) {
  predModWide <- predModWide %>%
    mutate(across(-patient, ~ c(scale(.))))
}

predModLong <- predModWide %>%
  pivot_longer(cols = -patient, names_to = "cellType", values_to = "proportion")


# Add patient characteristics
predSympWide <- predModWide %>%
  left_join(patientData, by = "patient")

predSympLong <- predModLong %>%
  left_join(patientData, by = "patient")
```

# 5.4 Statistical analysis

## 5.4.1 Simple statistical tests

```{r}
# Test for normality
resShapiro <- predSympLong %>%
  group_by(cellType) %>%
  summarise(pvalue = shapiro_test(proportion)$p.value) %>%
  add_significance("pvalue")

for(celltype in unique(predSympLong$cellType)) {
  subset <- predSympLong %>%
    filter(cellType == celltype)

  qqp <- ggpubr::ggqqplot(subset$proportion) +
    labs(title = paste("Q-Q plot for", celltype))

  print(qqp)
}

# Test for homogeneity of variance
resLevene <- predSympLong %>%
  levene_test(proportion ~ cellType)
```

```{r}
charCols <- c("Symptoms.2g", "Symptoms.4g", "Gender", "Macrophages.bin", "SMC.bin", "IPH.bin", "smc_macrophages_ratio", "plaquephenotype")

# Order cell types based on proportion to make sure that the significance bars will match the right boxplot
orderedCelltypes <- predSympLong %>%
  group_by(cellType) %>%
  summarise(proportion = mean(proportion)) %>%
  arrange(proportion) %>%
  pull(cellType)

predSympLong$cellType <- factor(predSympLong$cellType, levels = orderedCelltypes)

predSympLongFilter <- sympFilter(predSympLong, charCols)

resStats <- list()
resDunnList <- list()

for(symp in names(predSympLongFilter)) {
  df <- predSympLongFilter[[symp]]
  
  formula <- as.formula(sprintf("proportion ~ %s", symp))

  if(n_distinct(df[[symp]]) > 2) {
    res <- df %>%
      group_by(cellType) %>%
      kruskal_test(formula = formula) %>%
      add_significance("p")
    
    if(any(res$p <= 0.05)) {
      resDunn <- df %>%
        group_by(cellType) %>%
        dunn_test(formula = formula, p.adjust.method = "none") %>%
        add_significance("p") %>%
        add_xy_position(x = "cellType")
    
        resDunnList[[symp]] <- resDunn
      }
  } else {
    res <- df %>%
      group_by(cellType) %>%
      wilcox_test(formula = formula) %>% 
      add_significance("p") %>%
      add_xy_position(x = "cellType")
  }
  resStats[[symp]] <- res
}
```

## 5.4.2 General Linear Model

### 5.4.2.1 Fit GLM

```{r}
# Variables to be used in the fitting of the GLM
responseVar <- "Symptoms.2g" # MODIFY HERE

covariates <- c() # MODIFY HERE

cellTypes <- colnames(pred)[2:ncol(pred)]

# Generate formulas to be used
formulasGLM <- sapply(cellTypes, function(ct) {
  if(length(covariates) > 0) {
    as.formula(sprintf("%s ~ %s + %s", responseVar, ct, paste(covariates, collapse = " + ")))
  } else {
    as.formula(sprintf("%s ~ %s", responseVar, ct))
  }
})

# Fit GLM
resGLM <- lapply(formulasGLM, function(f) glm(f, family = binomial, data = predSympWide))

lapply(resGLM, function(r) summary(r))

# Extract results from all fitted models
tidyResGLM <- lapply(seq_along(resGLM), function(i) {
  tblRes <- tidy(resGLM[[i]], conf.int = TRUE) %>%
    mutate(signif = stars.pval(p.value),
           cellType = factor(names(resGLM)[i]))
})

tblAllResGLM <- do.call(rbind, tidyResGLM) %>%
  select(cellType, everything())

performanceResGLM <- lapply(resGLM, function(r) {
  tblGlance <- glance(r)
})

tblAllPerfGLM <- do.call(rbind, performanceResGLM)
```

### 5.4.2.2 Validate GLM

```{r}
# Plot residual plots
lapply(names(resGLM), function(name) {
  par(mfrow = c(2, 2))
  plot(resGLM[[name]], main = name)
})
```

### 5.4.2.3 Export GLM all results

```{r}

```

## 5.4.3 Cox Regression Model

```{r}
matCor <- cor(select(predModWide, -patient))

ggcorrplot::ggcorrplot(matCor, hc.order = TRUE, type = "lower", lab = TRUE)
```

### 5.4.3.1 Fit Cox proportional hazards model

```{r}
# Variables to be used in the Cox regression
timeVar <- "EP_major_time" # MODIFY HERE

statusVar <- "EP_major" # MODIFY HERE

covariates <- c("Gender", "Age") # MODIFY HERE

cellTypes <- colnames(pred)[2:ncol(pred)]

# Generate formulas to be used
formulasCox <- sapply(cellTypes, function(ct) {
  if(length(covariates) > 0) {
    as.formula(sprintf("Surv(%s, %s) ~ %s + %s", timeVar, statusVar, ct, paste(covariates, collapse = " + ")))
  } else {
    as.formula(sprintf("Surv(%s, %s) ~ %s", timeVar, statusVar, ct))
  }
})

# Fit Cox proportional hazards model
resCox <- lapply(formulasCox, function(f) {
  coxModel <- tryCatch(coxph(f, data = predSympWide), 
                       error = function(e) {
                         message("Cox regression failed for formula: ", deparse(f))
                       })
})

lapply(resCox, function(r) summary(r))

# Extract results from all fitted models
tidyResCox <- lapply(seq_along(resCox), function(i) {
  tblRes <- tidy(resCox[[i]], conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(signif = stars.pval(p.value),
           cellType = factor(names(resCox)[i]))
})

tblAllResCox <- do.call(rbind, tidyResCox) %>%
  select(cellType, everything()) %>%
  rename(HR = estimate)

performanceResCox <- lapply(resCox, function(x) {
  tblGlance <- glance(x)
})

tblAllPerfCox <- do.call(rbind, performanceResCox)
```

### 5.4.3.2 Validate Cox regression model

```{r}
testCoxAssump <- lapply(resCox, function(r) cox.zph(r))
print(testCoxAssump)

pTestCoxAssump <- lapply(testCoxAssump, function(t) ggcoxzph(t))
invisible(lapply(pTestCoxAssump, print))
```

```{r}
lapply(resCox, function(r) ggcoxdiagnostics(r, type = "martingale") + labs(title = r))
```

```{r}
lapply(resCox, function(r) ggcoxdiagnostics(r, type = "deviance") + labs(title = r))
```

```{r}
lapply(resCox, function(r) ggcoxdiagnostics(r, type = "dfbeta"))
```

```{r}
lapply(resCox, function(r) ggcoxdiagnostics(r, type = "scaledsch"))
```

```{r}
lapply(resCox, function(r) ggcoxfunctional(r))
```

### 5.4.3.3 Export Cox regression results

```{r}

```

# 5.5 Graphs

## 5.5.1 Boxplot of predicted cellular proportion, divided on patient characteristics

```{r}
labelSymp <- list(Symptoms.2g = c("mild", "severe"),
                  Symptoms.4g = c("asymptomatic", "ocular", "tia", "stroke"),
                  Gender = c("female", "male"),
                  Macrophages.bin = c("no/minor", "moderate/heavy"),
                  SMC.bin = c("no/minor", "moderate/heavy"),
                  IPH.bin = c("no", "yes"),
                  smc_macrophages_ratio = c("smc dominant", "equal", "macrophages dominant"),
                  plaquephenotype = c("fibrous", "fibroatheromatous", "atheromatous")) # MODIFY HERE

listFilterPredSympLong <- list()

for(symp in names(labelSymp)) {
  filterPredSympLong <- predSympLong %>%
    filter(!is.na(!!sym(symp)) & !!sym(symp) != -888 & !!sym(symp) != -999) %>%
    select(patient, cellType, proportion, !!sym(symp))
  
  listFilterPredSympLong[[symp]] <- filterPredSympLong
}
```

### 5.5.1.1 Without significance bars

```{r}
plotList <- list()

for(symp in names(labelSymp)) {
  nLabels <- as.character(length(labelSymp[[symp]]))
  
  pSymp <- ggplot(listFilterPredSympLong[[symp]], aes(x = reorder(cellType, proportion), y = proportion)) +
  geom_boxplot(aes(fill = as.factor(!!sym(symp))), position = position_dodge(width = 0.8), width = 0.5, linewidth = 0.3, outlier.size = 1, outlier.shape = 1) +
  labs(title = pattern,
       x = "",
       y = "Estimated cell type proportion",
       fill = symp) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     breaks = seq(0, 1, by = 0.1)) +
  scale_fill_manual(values = blueColorPal[[nLabels]], labels = labelSymp[[symp]]) + # MODIFY HERE; custom color palette options in functions_figures_R.R
  theme_classic(base_family = "Helvetica") +
  boxplotTheme()
  
  plotList[[symp]] <- pSymp
}

print(plotList)
```

```{r}
path <- paste0("output/plots/boxplot_cell_type_proportion_symptoms/", pattern)

createDir(path)

for(name in names(plotList)) {
  tiff(paste0(path, "/", name, "_", pattern, ".tiff"))
  
  print(plotList[[name]])
  
  dev.off()
}
```

### 5.5.1.2 With significance bars from Section 5.4.1

```{r}
signPlotList <- list()

for(symp in names(resStats)) {
  if(any(resStats[[symp]]$p.signif != "ns")) {
    if(symp %in% names(resDunnList)) {
      df <- resDunnList[[symp]]
    } else {
      df <- resStats[[symp]]
    }
    
    pAdd <- plotList[[symp]]
    
    pSign <- pAdd +
      stat_pvalue_manual(df, label = "p.signif", hide.ns = TRUE, tip.length = 0.5)
    
    signPlotList[[symp]] <- pSign
  }
}

print(signPlotList)
```

```{r}
pathSign <- paste0("output/plots/boxplot_cell_type_proportion_symptoms_significant/", pattern)

if(length(signPlotList) != 0) {
  createDir(pathSign)
  for(name in names(signPlotList)) {
    tiff(paste0(pathSign, "/", name, "_", pattern, ".tiff"))
    
    print(signPlotList[[name]])
    
    dev.off()
    }
  } else {
    cat("List is empty")
}
```

## 5.5.2 Forest plot

### 5.5.2.1 GLM

```{r}
pGLM <- ggplot(allGLM, aes(x = cellType, y = beta)) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper, color = sig), 
                width = 0.2, size = 1, linewidth = 0.6) +
  geom_point(aes(fill = sig), shape = 21, size = 2, stroke = NA) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = 0.2) +
  coord_flip() +
  labs(x = "",
       y = expression(beta ~ "coefficient (" * bold(symptomatic) * ")"),
       color = expression(bold(significant))) +
  scale_color_manual(values = c("#7C7C7C", "#00868B")) +
  scale_fill_manual(values = c("black", "#00868B")) +
  # scale_y_continuous(limits = c(-0.6, 0.5),
  #                    breaks = seq(-0.5, 0.5, 0.25),
  #                    labels = c("-0.5", "-0.25", "0", "0.25", "0.5")) +
  guides(color = guide_legend(override.aes = list(shape = NA, linewidth = 0.6)),
         fill = "none") +
  theme_bw(base_family = "Helvetica") +
  theme(panel.border = element_rect(linewidth = 0.4, color = "black"),
        panel.grid.major = element_line(linewidth = 0.2),
        axis.ticks = element_line(linewidth = 0.2),
        axis.ticks.length = unit(0.04, "cm"),
        axis.title = element_text(size = 6),
        axis.text = element_text(size = 6),
        legend.margin = margin(l = -3),
        legend.title = element_text(size = 6, margin = margin(l = -6)),
        legend.text = element_text(size = 6),
        legend.key.size = unit(0.26, "cm"),
        aspect.ratio = 1)

print(pGLM)

ggsave(paste0("output/plots/glm_forest_plot/tiff/scaden_", pattern, "_glm_gender.tiff", collapse = ""), 
     width = 15, height = 15, units = "cm", dpi = 300)

ggsave(paste0("output/plots/glm_forest_plot/svg/scaden_", pattern, "_glm_gender.svg", collapse = ""),
     width = 15, height = 15, units = "cm", dpi = 300)
```

### 5.5.2.2 Cox regression model

```{r}
# https://rpubs.com/mbounthavong/forest_plots_r
# Create the base plot for the different text columns
baseCol <- ggplot(dfAllResCox, aes(y = cellType)) +
  theme(text = element_text(family = "Helvetica"),
        plot.title = element_text(size = 9, face = "bold", vjust = -1),
        axis.text.x = element_text(color = "white", hjust = -3),
        axis.line = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks = element_blank(),
        axis.title.y = element_blank(), 
        legend.position = "none",
        panel.background = element_blank(), 
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.background = element_blank())

pCellType <- baseCol +
  geom_text(aes(x = -Inf, label = cellType), hjust = 0, size = 2.5, family = "Helvetica") +
  ggtitle("Cell type") +
  annotate("text", x = 0, y = Inf, label = "") +
  theme(plot.title = element_text(hjust = 0))

pForest <- ggplot(dfAllResCox, aes(x = cellType, y = HR)) +
  geom_hline(yintercept = 1, linewidth = 0.3, color = "#878787") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper, color = Sig.), width = NA, linewidth = 0.3) +
  geom_point(aes(color = Sig.), shape = 22, size = 1, stroke = 0.8) +
  geom_point(aes(fill = Sig.), shape = 22, size = 1.2, stroke = NA) + 
  coord_flip() +
  labs(x = "",
       y = expression(bold("Hazard Ratio")),
       color = expression(bold(significant))) +
  scale_color_manual(values = c("#878787", "#4848F7")) + # Colors for the outlines
  scale_fill_manual(values = c("black", "#FF0000")) + # Colors for the squares
  # scale_y_continuous(limits = c(0.5, 1.5),
                     # breaks = seq(0.5, 1.5, 0.5),
                     # labels = c("0.5", "1", "1.5")) +
  guides(color = "none",
         fill = "none",
         x = guide_axis(cap = "both")) +
  theme_classic() +
  theme(axis.text.x = element_text(family = "Helvetica", size = 7),
        axis.title.x = element_text(family = "Helvetica", size = 7),
        axis.line.y = element_blank(),
        axis.ticks.y= element_blank(),
        axis.text.y= element_blank(),
        axis.title.y= element_blank(),
        axis.ticks.length.x = unit(0.2, "cm"),
        axis.line = element_line(color = "#878787", linewidth = 0.3),
        axis.ticks = element_line(color = "#878787", linewidth = 0.3),
        panel.background = element_blank(),
        plot.background = element_blank())

pHR <- baseCol +
  geom_text(aes(x = 0, label = `HR [95%CI]`), hjust = 0.5, size = 2.5, family = "Helvetica") +
  ggtitle("HR [95% CI]") +
  theme(plot.title = element_text(hjust = 0.5))

pPValue <- baseCol +
  geom_text(aes(x = 0, label = p), hjust = 0.5, size = 2.5, family = "Helvetica") +
  ggtitle("p") +
  theme(plot.title = element_text(hjust = 0.5))

pSig <- baseCol +
  geom_text(aes(x = 0, label = `Sig.`), hjust = 0.5, size = 2.5, family = "Helvetica") +
  ggtitle("Sig.") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
pCellType + pForest + pHR + pPValue + pSig + plot_layout(design = c(area(t = 0, l = 0, b = 5, r = 39),
                                                                    area(t = 0, l = 36, b = 5, r = 57),
                                                                    area(t = 0, l = 57, b = 5, r = 72),
                                                                    area(t = 0, l = 69, b = 5, r = 80),
                                                                    area(t = 0, l = 80, b = 5, r = 80)))

ggsave(paste0("output/plots/cox_forest_plot/tiff/scaden_", pattern, "_cox.tiff", collapse = ""), 
     width = 15, height = 15, units = "cm", dpi = 300)

ggsave(paste0("output/plots/cox_forest_plot/svg/scaden_", pattern, "_cox.svg", collapse = ""),
     width = 15, height = 15, units = "cm", dpi = 300)
```

# 5.6 Session info

```{r}
devtools::session_info()
```
